`r if(knitr:::is_latex_output()) '\\appendix'`

`r if(!knitr:::is_latex_output()) '# (APPENDIX) Appendix {-}'`

<!--
If you feel it necessary to include an appendix, it goes here.
-->


# Appendix

## Survival Analysis 

```{r more-hists, echo=F, cache=T, fig.cap="Histograms of Relationship Duration by Censored Status & Type"}
library(tidyverse)
dataset <- readRDS("~/nsfg_data_cleaning/Objects/altersegos_survdat.rds")

dataset$reltype2 = ifelse(dataset$reltype %in% "Other", 1, 2)
dataset$len <- dataset$t_c-dataset$t_o
dataset$reltype4[dataset$len<=1 & dataset$reltype2==1] <- 1
dataset$reltype4[dataset$len>1 & dataset$reltype2==1] <- 2
dataset$reltype4[dataset$reltype=="Cur/Fmr Cohab"] <- 3
dataset$reltype4[dataset$reltype=="Cur/Fmr Spouse"] <- 4

dat <- dataset %>% mutate(reltype = ifelse(reltype %in% "Cur/Fmr Cohab", "Cohab",
                                       ifelse(reltype %in% "Cur/Fmr Spouse", "Marriage",
                                        reltype)),
                      reltype2 = ifelse(reltype %in% "Cohab", "Cohab/Marriage",
                                       ifelse(reltype %in% "Marriage", "Cohab/Marriage",
                                        ifelse(reltype %in% "Other", "Casual", reltype))),
                      reltype4 = ifelse(reltype4==1, "Casual-short",
                                        ifelse(reltype4==2, "Casual-long",
                                               ifelse(reltype4==3, "Cohab",
                                                      ifelse(reltype4==4, "Marriage", NA)))),
                      censorstatus = ifelse(censored==1, "Completed", "Ongoing"))


 dat %>%
     ggplot(aes(x = t_c_years, fill = reltype2, weights = e.weight)) +
     geom_histogram(binwidth = 0.25) +
     scale_x_continuous(name="Relationship Age, Years") +
     scale_y_continuous(name="Weighted Count in Millions") +
     theme(aspect.ratio=1) +
     labs(fill="Rel Type") +
     scale_fill_manual(name="Rel Type", values=c("#014d64", "#6974a7")) +
     facet_wrap(~reltype2+censorstatus) +
    theme(legend.position="none")
```

## Chlamydia & EPT

```{r uncalibrated-networks, echo=F, message=F, warning=F}
source("~/Documents/Dissertation/R/EPT/functions/ddaf_functions.R")
source("~/Documents/Dissertation/R/Duration/Functions/as_nfsg_rels.R")
library(EpiModel)
library(ggplot2)
library(srvyr)
library(kableExtra)
library(tidyverse)

sim <- readRDS("~/Documents/Dissertation/R/EPT/calibration/uncalibrated_sim.rds")

mdist1 <- get_dist(sim, network=1, ndynamic=2)
#mdist1[[1]];mdist1[[2]]

odist1 <- get_dist(sim, network=2, ndynamic=2)
#odist1[[1]];odist1[[2]]
```

```{r coefs,  message=F, warning=F, echo=F, cache=T, out.width="70%"}
library(broom)
library(here)
library(tidyverse)
library(kableExtra)
fit.other <- readRDS(here("data", "ept", "fit.other.rds"))
fit.marcoh <- readRDS(here("data", "ept", "fit.marcoh.rds"))
fit.once  <- readRDS(here("data", "ept", "fit.once.rds"))
netest <- readRDS(here("data", "ept", "calibrated_netests.rds"))

# update calibrated coefs
fit.marcoh$coef[3:15] <- netest[[1]]$coef.form[2:14]
fit.other$coef[3:30] <- netest[[2]]$coef.form[2:29]


fits <- rbind(tidy(fit.marcoh), tidy(fit.other), tidy(fit.once))

durs <- data.frame(Network= c("Marriage", "Casual", "Inst"), `Duration` = c(476, 55, 1))


kable(fits[1:15,], booktabs=T, col.names = c("Model Term", "Estimate", "SE", "Statistic", "Pvalue"),
      caption="Marriage/Cohabitation Network: Summary of Formation Model Fit ") %>%
  kable_styling()

kable(fits[16:45,], booktabs=T, col.names = c("Model Term", "Estimate", "SE", "Statistic", "Pvalue"),
      caption="Casual Network: Summary of Formation Model Fit ") %>%
  kable_styling()

kable(fits[46:56,], booktabs=T, col.names = c("Model Term", "Estimate", "SE", "Statistic", "Pvalue"),
      caption="Instantaneous: Summary of Formation Model Fit ") %>%
  kable_styling()

#kable(durs, booktabs=T, caption="Mean Relationship Duration in Weeks") %>%
#  kable_styling()
```

```{r egodata-prep, echo=FALSE, fig.align='center', out.width="60%", fig.cap="Mean Degree by Ego Age and Relationship Type.", cache=T}
egoplotdat <- cbind(egodat[[2]][,1:2], egodat[[4]][,2])
names(egoplotdat) <- c("Ego Age", "Marriage/Cohab", "Casual")
egoplotdat <-  gather(egoplotdat, Type, `Mean Degree`, -`Ego Age`)
ggplot() +
  geom_point(data=egoplotdat, aes(x=`Ego Age`, y=`Mean Degree`, color=Type)) +
  scale_color_manual(values=c("darkblue", "darkred")) +
  theme(legend.position = c(0.15, 0.8))
```

include plots for the other terms - absdiff(sqrtage), concurrent, debuted etc  

several general trends in relationship formation (finish write-up and cite) --

-	individuals often select partners that are not their exact age
-	this difference in partner ages often increases over the life course (i.e. adults usually have wider age differences between their partners than do adolescents)
-	it is common for men to partner with younger women (although the sex differences in relationship formation are not explored in this model, it’s important to note that in a more realistic model the effect of aging out would disproportionately affect the women whose partners age out before them

(include model terms and coefs and explain terms)
(full description of EpiModelHIV module flow w/ parameters in appendix, brief overview here)

Cross network terms - we're going to avoid them due to complications  



